Towards Proof-Reconstruction of Problems in Classical Propositional Logic

### 2017-10-01 paper/sections/proof-reconstruction.tex
[ paper ] canonicalize lemmas and theorems.

Then, we solve the termination problem by modifying problematic functions
to be \emph{structural recursive}~\cite{Coquand1992,Abel2002,Bove2005}

@@

\begin{align}
\begin{split}
\label{eq:unshunt-definition}
&\unshunt_{0}(φ) =\\
&\hspace{3mm}\begin{cases}
\unshunt_{0}((φ₁ ∧ φ₂) ⇒ φ₃),
  & \text{if }φ ≡ φ₁ ⇒ (φ₂ ⇒ φ₃)\\
\unshunt_{0}(φ₁ ⇒ φ₂) ∧ \unshunt_{0}(φ₁ ⇒ φ₃),
  & \text{if }φ ≡ φ₁ ⇒ (φ₂ ∧ φ₃)\\
φ, &\text{otherwise.}
\end{cases}
\end{split}
\end{align}

@@

Thus, we define the function
$\unshunt_{1} : \Prop \to \Nat \to \Prop$
by structural recursion in its second argument in
Eq.~\ref{eq:unshunt-structured}.

\begin{align}
\label{eq:unshunt-structured}
\begin{split}
&\unshunt_{1}(φ, \zero)\hspace{3mm}= φ\\
&\unshunt_{1}(φ, \suc(n)) =\\
&\hspace{3mm}\begin{cases}
\unshunt_{1}((φ₁ ∧ φ₂) ⇒ φ₃, n),
  &\text{if } φ ≡ φ₁ ⇒ (φ₂ ⇒ φ₃)\\
\unshunt_{1}(φ₁ ⇒ φ₂, n) ∧\ \unshunt_{1}(φ₁ ⇒ φ₃, n),
   &\text{if } φ ≡ φ₁ ⇒ (φ₂ ∧ φ₃).
\end{cases}
\end{split}
\end{align}

As we mentioned above, the natural number in the second argument of the function used in Eq.~\ref{eq:unshunt-structured} stands for a complexity measure of the function, in this case $\unshunt_{1}$ function.

@@

Recall, we want to maintain the same outputs as the original \unshunt does in Eq.~\ref{eq:unshunt-definition} and the complexity measure of it, is a bound for the number of times this function can call itself.

Therefore, we define an appropriate complexity measure
$\unshunt_{cm}: \Prop → \NN$ for our $\unshunt_{1}$ function presented in
Eq.~\ref{eq:unshunt-rank-definition}.

\begin{equation}
\label{eq:unshunt-rank-definition}
\unshunt_{cm}(φ)=
\begin{cases}
\unshunt_{cm}(φ₃) + 2, &\text{if }φ ≡ φ₁ ⇒ (φ₂ ⇒ φ₃)\\
\max(\unshunt_{cm}(φ₂),\unshunt_{cm}(φ₃)) + 1,
  &\text{if }  φ ≡ φ₁ ⇒ (φ₂ ∧ φ₃)\\
0, &\text{otherwise.}
\end{cases}
\end{equation}

Finally, the curated version of $\unshunt_{0}$ function is presented
in Eq.~\ref{eq:unshunt-fixed}.

\begin{equation}
\label{eq:unshunt-fixed}
\unshunt(\varphi) := \unshunt_{1}(\varphi, \unshunt_{cm}(\varphi)).
\end{equation}

@@

In the following, we show how Theorem.~\ref{thm:thm-strip} proves the correctness of the process mentioned above. But first, let prove some lemmas for $\unshunt$ function presented in Eq.~\ref{eq:unshunt-fixed},
and $split$ function from the definition in Eq.~\ref{eq:split-definition}.

\begin{lemma}[\thmunshunt]
  \label{lem:unshunt}
  $Γ ⊢ \unshunt(φ)$, then $Γ ⊢ φ$.
\end{lemma}

\begin{proof} Use induction on the cases defined by the outcome of the
$\unshunt$ function in Eq.~\ref{eq:unshunt-structured}.
When $n = 0$, by definition, $Γ ⊢ φ$.
When $n = 1$, we apply the following theorem from~\cite{AgdaProp}

@@

(φ₁ ⇒ (φ₂ ⇒ φ₃))
\begin{equation*}
  \begin{bprooftree}
  \AxiomC{$Γ ⊢ \unshunt(φ₁ ⇒ (φ₂ ⇒ φ₃), n)$}
  \RightLabel{By def.}
  \UnaryInfC{$Γ ⊢ \unshunt(φ₁ ⇒ (φ₂ ⇒ φ₃), n-1)$}
  \RightLabel{\thmunshunt}
  \UnaryInfC{Γ ⊢ (φ₁ ∧ φ₂) ⇒ φ₃}
  \RightLabel{∧⇒-to-⇒⇒}
  \UnaryInfC{Γ ⊢ φ₁ ⇒ (φ₂ ⇒ φ₃).}
  \end{bprooftree}
\end{equation*}

@@

($φ₁ ⇒ (φ₂ ∧ φ₃)$)
\begin{equation*}
\scalebox{0.95}{
($\mathcal{D}_1$)
\begin{bprooftree}
\AxiomC{$Γ ⊢ \unshunt(φ₁ ⇒ φ₂, n) ∧ \unshunt(φ₁ ⇒ φ₃, n)$}
\RightLabel{∧-proj₁}
\UnaryInfC{$Γ ⊢ \unshunt(φ₁ ⇒ φ₂, n)$}
\RightLabel{By def.}
\UnaryInfC{$Γ ⊢ \unshunt(φ₁ ⇒ φ₂, n-1)$}
\RightLabel{\thmunshunt}
\UnaryInfC{$Γ ⊢ φ₁ ⇒ φ₂$,}
\end{bprooftree}
}
\end{equation*}

\begin{equation*}
(\mathcal{D}_2)
\begin{bprooftree}
\AxiomC{$Γ ⊢ \unshunt(φ₁ ⇒ φ₂) ∧ \unshunt(φ₁ ⇒ φ₃)$}
\RightLabel{∧-proj₂}
\UnaryInfC{$Γ ⊢ \unshunt(φ₁ ⇒ φ₃, n)$}
\RightLabel{By def.}
\UnaryInfC{$Γ ⊢ \unshunt(φ₁ ⇒ φ₃, n-1)$}
\RightLabel{\thmunshunt}
\UnaryInfC{$Γ ⊢ φ₁ ⇒ φ₃$,}
\end{bprooftree}
\end{equation*}

@@

\begin{equation}
\label{eq:split-definition}
\mysplit(φ) =
\begin{cases}
\unshunt(\mysplit(φ₁)) &\text{if }φ ≡ φ₁ ∧ φ₂\\
∧\ \unshunt(φ₁ ⇒ \mysplit(φ₂)), &\\[2.5mm]
\unshunt(¬ φ₁ ⇒ \mysplit(φ₂)),
  & \text{if }φ ≡ φ₁ ∨ φ₂\\[2.5mm]
\unshunt(φ₁ ⇒ \mysplit(φ₂)),
  & \text{if }φ ≡ φ₁ ⇒ φ₂\\[2.5mm]
\unshunt(φ₁ ⇒ \mysplit(φ₂))
  & \text{if }φ ≡ φ₁ ⇔ φ₂\\
∧\ \unshunt(φ₂ ⇒ \mysplit(φ₁)),
  &\\[2.5mm]
\unshunt(φ₁ ⇒ \mysplit(¬ φ₂)),
  & \text{if }φ ≡ ¬ (φ₁ ∧ φ₂)\\[2.5mm]
\unshunt(\mysplit(¬ φ₁))
  & \text{if }φ ≡ ¬ (φ₁ ∨ φ₂)\\
∧\ \unshunt(¬ φ₁ ⇒ \mysplit(¬ φ₂)),
  &\\[2.5mm]
\unshunt(\mysplit(φ₁))
  & \text{if }φ ≡ ¬ (φ₁ ⇒ φ₂)\\
∧\ \unshunt(φ₁ ⇒ \mysplit(¬ φ₂)),
  &\\[2.5mm]
\unshunt(φ₁ ⇒ \mysplit(¬ φ₂))
  & \text{if }φ ≡ ¬ (φ₁ ⇔ φ₂)\\
∧\ \unshunt(¬ φ₂ ⇒ \mysplit(φ₁)),
  &\\[2.5mm]
\unshunt(\mysplit(φ₁)),
  & \text{if }φ ≡ ¬ ¬ φ₁\\
⊤,
  & \text{if }φ ≡ ¬ ⊥\\
⊥,
  & \text{if }φ ≡ ¬ ⊤.
\end{cases}
\end{equation}

@@

\begin{equation*}
(\mathcal{D})\hspace{3mm}
  \begin{bprooftree}
  \AxiomC{$Γ ⊢ \mysplit(φ,n)$}
  \RightLabel{By def.}
  \UnaryInfC{$Γ ⊢ \unshunt(φ₁) ∧ \unshunt(φ₁ ⇒ \mysplit(φ₂,n-1))$}
  \RightLabel{∧-proj₁}
  \UnaryInfC{$Γ ⊢ \unshunt(φ₁)$}
  \RightLabel{\thmunshunt}
  \UnaryInfC{$Γ ⊢ φ₁$.}
  \end{bprooftree}
\end{equation*}

\begin{equation*}
  \begin{bprooftree}
  \AxiomC{$\mathcal{D}$}
  \UnaryInfC{Γ ⊢ φ₁}

  \AxiomC{$Γ ⊢ \mysplit(φ,n)$}
  \RightLabel{By def.}
  \UnaryInfC{$Γ ⊢ \unshunt(φ₁) ∧ \unshunt(φ₁ ⇒ \mysplit(φ₂,n-1))$}
  \RightLabel{$∧$-proj$₂$}
  \UnaryInfC{$Γ ⊢ \unshunt(φ₁ ⇒ \mysplit(φ₂,n-1))$}
  \RightLabel{\thmunshunt}
  \UnaryInfC{$Γ ⊢ φ₁ ⇒ \mysplit(φ₂,n-1)$}
  \RightLabel{$⇒$-elim}
  \BinaryInfC{$Γ ⊢ \mysplit(φ₂,n-1)$}
  \RightLabel{\thmsplit}
  \UnaryInfC{$Γ ⊢ φ₂.$}
  \end{bprooftree}
\end{equation*}

@@

($φ ≡ φ₁ ⇒ φ₂$).
\begin{equation*}
  \begin{bprooftree}
  \AxiomC{}
  \RightLabel{assume φ₁}
  \UnaryInfC{$Γ , φ₁ ⊢ φ₁$}
  \AxiomC{$Γ ⊢ \mysplit(φ,n-1)$}
  \RightLabel{By def.}
  \UnaryInfC{$Γ ⊢ \unshunt(φ₁ ⇒ \mysplit(φ₂,n-1))$}
  \RightLabel{\thmunshunt}
  \UnaryInfC{$Γ ⊢ φ₁ ⇒ \mysplit(φ₂,n-1)$}
  \RightLabel{weaken φ₁}
  \UnaryInfC{$Γ , φ₁ ⊢ φ₁ ⇒ \mysplit(φ₂,n-1)$}
  \RightLabel{⇒-elim}
  \BinaryInfC{$Γ , φ₁ ⊢ \mysplit(φ₂,n-1)$}
  \RightLabel{\thmsplit}
  \UnaryInfC{$Γ , φ₁ ⊢ φ₂$}
  \RightLabel{⇒-intro}
  \UnaryInfC{$Γ ⊢ φ₁ ⇒ φ₂$.}
  \end{bprooftree}
\end{equation*}
\vskip 3mm

@@

\begin{equation*}
(\mathcal{D})\hspace{3mm}
\begin{bprooftree}
\AxiomC{$Γ ⊢ \mysplit(φ, n)$}
\RightLabel{By def.}
\UnaryInfC{$Γ ⊢ \unshunt(¬ φ₁) ∧ \unshunt(¬ φ₁ ⇒ \mysplit(φ₂, n))$}
\RightLabel{∧-proj₁}
\UnaryInfC{$Γ ⊢ \unshunt(¬ φ₁)$}
\RightLabel{\thmunshunt}
\UnaryInfC{$Γ ⊢ ¬ φ₁$,}
\end{bprooftree}
\end{equation*}

@@

  \AxiomC{$Γ ⊢ \mysplit(φ,n-1)$}
  \RightLabel{By def.}
  \UnaryInfC{$Γ ⊢ \unshunt(¬ φ₁) ∧ \unshunt(¬ φ₁ ⇒ \mysplit(φ₂,n-1))$}
  \RightLabel{∧-proj₂}
  \UnaryInfC{$Γ ⊢ \unshunt(¬ φ₁ ⇒ \mysplit(φ₂,n-1))$}
  \RightLabel{\thmunshunt}
  \UnaryInfC{$Γ ⊢ ¬ φ₁ ⇒ \mysplit(φ₂,n-1)$}
  \RightLabel{⇒-elim}
  \BinaryInfC{$Γ ⊢ \mysplit(φ₂,n-1)$}
  \RightLabel{\thmsplit}
  \UnaryInfC{$Γ ⊢ ¬ φ₂.$}
  \end{bprooftree}
\end{equation*}
\end{proof}

@@

\begin{equation*}
  \begin{bprooftree}
  \AxiomC{Γ ⊢ q}
  \RightLabel{∨-intro₂}
  \UnaryInfC{Γ ⊢ p ∨ q}
  \RightLabel{∨-intro₁}
  \UnaryInfC{Γ ⊢ (p ∨ q) ∨ r}

  \AxiomC{Γ ⊢ p}
  \RightLabel{∨-intro₁}
  \UnaryInfC{Γ ⊢ p ∨ q}
  \RightLabel{∨-intro₁}
  \UnaryInfC{Γ ⊢ (p ∨ q) ∨ r}

  \RightLabel{∨-elim}
  \BinaryInfC{Γ, q ∨ p ⊢ (p ∨ q) ∨ r}
  \end{bprooftree}
\end{equation*}

\begin{equation*}
  \begin{bprooftree}
  \AxiomC{Γ ⊢ r}
  \RightLabel{∨-intro₂}
  \UnaryInfC{Γ ⊢ (p ∨ q) ∨ r}

  \AxiomC{Γ, q ∨ p ⊢ (p ∨ q) ∨ r}

  \RightLabel{∨-elim}
  \BinaryInfC{Γ, r ∨ (q ∨ p) ⊢ (p ∨ q) ∨ r}

  \RightLabel{⇒-intro}
  \UnaryInfC{Γ ⊢ r ∨ (q ∨ p) ⇒ (p ∨ q) ∨ r}

@@

In a disjunction, $φ = φ₁ ∨ φ₂ ∨ \cdots ∨ φₙ$, we say $ψ ∈_{∨} φ$,
if there is some $i = 1, \cdots, n$ such that $ψ = φᵢ$.
Then, we define the $rm_{∨}$ function to remove in a disjunction φ, those propositions ψ repeated in the formula. The formula $φ$ must be a right-associative disjunction.

@@

Now, we have removed redundancies in the disjunctions, we move forward and we remove redundancies at the level of the conjunction of the CNFs by applying the $rm_{∧}$ function, we remove conjunct repeated in the formula. We define $ψ ∈_{∧} φ$ similar as we described above for disjunction. In a conjunction, $φ = φ₁ ∧ φ₂ ∧ \cdots ∧ φₙ$, we say
$ψ ∈_{∧} φ$, if there is some $i = 1, \cdots, n$ such that $ψ = φᵢ$.

@@

% subsection normalization (end)
% --------------------------------------------------------------------

>>>>

Then, we solve the termination problem by modifying problematic functions
to be \emph{structural recursive}~\cite{Coquand1992,Abel2002,Bove2005}.

@@

\begin{align}
\begin{split}
\label{eq:unshunt-definition}
&unshunt_{0}(φ) =\\
&\hspace{3mm}\begin{cases}
unshunt_{0}((φ₁ ∧ φ₂) ⇒ φ₃),
  & \text{if }φ ≡ φ₁ ⇒ (φ₂ ⇒ φ₃)\\
unshunt_{0}(φ₁ ⇒ φ₂) ∧ unshunt_{0}(φ₁ ⇒ φ₃),
  & \text{if }φ ≡ φ₁ ⇒ (φ₂ ∧ φ₃)\\
φ, &\text{otherwise.}
\end{cases}
\end{split}
\end{align}

@@

Thus, we define the function
$unshunt_{1} : \Prop \to \Nat \to \Prop$
by structural recursion in its second argument in
Eq.~\ref{eq:unshunt-structured}.

\begin{align}
\label{eq:unshunt-structured}
\begin{split}
&unshunt_{1}(φ, \zero)\hspace{3mm}= φ\\
&unshunt_{1}(φ, \suc(n)) =\\
&\hspace{3mm}\begin{cases}
unshunt_{1}((φ₁ ∧ φ₂) ⇒ φ₃, n),
  &\text{if } φ ≡ φ₁ ⇒ (φ₂ ⇒ φ₃)\\
unshunt_{1}(φ₁ ⇒ φ₂, n) ∧\ unshunt_{1}(φ₁ ⇒ φ₃, n),
   &\text{if } φ ≡ φ₁ ⇒ (φ₂ ∧ φ₃).
\end{cases}
\end{split}
\end{align}

As we mentioned above, the natural number in the second argument of the
function used in Eq.~\ref{eq:unshunt-structured} stands for a complexity
measure of the function, in this case $unshunt_{1}$ function.

@@

Recall, we want to maintain the same outputs as the original unshunt
does in Eq.~\ref{eq:unshunt-definition} and the complexity measure of it,
is a bound for the number of times this function can call itself.

Therefore, we define an appropriate complexity measure
$unshunt_{cm}: \Prop → \NN$ for our $unshunt_{1}$ function presented in
Eq.~\ref{eq:unshunt-rank-definition}.

\begin{align}
\begin{split}
\label{eq:unshunt-rank-definition}
&unshunt_{cm}(φ) =\\
&\hspace{3mm}
\begin{cases}
unshunt_{cm}(φ₃) + 2, &\text{if }φ ≡ φ₁ ⇒ (φ₂ ⇒ φ₃)\\
\max(unshunt_{cm}(φ₂),unshunt_{cm}(φ₃)) + 1,
  &\text{if }  φ ≡ φ₁ ⇒ (φ₂ ∧ φ₃)\\
0, &\text{otherwise.}
\end{cases}
\end{split}
\end{align}

Finally, the curated version of $unshunt_{0}$ function is presented
in Eq.~\ref{eq:unshunt-fixed}.

\begin{equation}
\label{eq:unshunt-fixed}
unshunt(\varphi) := unshunt_{1}(\varphi, unshunt_{cm}(\varphi)).
\end{equation}

@@

In the following, we show how Theorem.~\ref{thm:thm-strip} proves the correctness of the process mentioned above. But first, let prove some lemmas for $unshunt$ function presented in Eq.~\ref{eq:unshunt-fixed},
and $split$ function from the definition in Eq.~\ref{eq:split-definition}.

\begin{lemma}[lem-unshunt]
  \label{lem:unshunt}
  $Γ ⊢ unshunt(φ)$, then $Γ ⊢ φ$.
\end{lemma}

\begin{proof} Use induction on the cases defined by the outcome of the
$unshunt$ function in Eq.~\ref{eq:unshunt-structured}.
When $n = 0$, by definition, $Γ ⊢ φ$.
When $n = 1$, we apply the following theorem from~\cite{AgdaProp}

@@

(φ₁ ⇒ (φ₂ ⇒ φ₃))
\begin{equation*}
  \begin{bprooftree}
  \AxiomC{$Γ ⊢ unshunt(φ₁ ⇒ (φ₂ ⇒ φ₃), n)$}
  \RightLabel{By def.}
  \UnaryInfC{$Γ ⊢ unshunt(φ₁ ⇒ (φ₂ ⇒ φ₃), n-1)$}
  \RightLabel{lem-unshunt}
  \UnaryInfC{Γ ⊢ (φ₁ ∧ φ₂) ⇒ φ₃}
  \RightLabel{∧⇒-to-⇒⇒}
  \UnaryInfC{Γ ⊢ φ₁ ⇒ (φ₂ ⇒ φ₃).}
  \end{bprooftree}
\end{equation*}

@@

($φ₁ ⇒ (φ₂ ∧ φ₃)$)
\begin{equation*}
\scalebox{0.95}{
($\mathcal{D}_1$)
\begin{bprooftree}
\AxiomC{$Γ ⊢ unshunt(φ₁ ⇒ φ₂, n) ∧ unshunt(φ₁ ⇒ φ₃, n)$}
\RightLabel{∧-proj₁}
\UnaryInfC{$Γ ⊢ unshunt(φ₁ ⇒ φ₂, n)$}
\RightLabel{By def.}
\UnaryInfC{$Γ ⊢ unshunt(φ₁ ⇒ φ₂, n-1)$}
\RightLabel{lem-unshunt}
\UnaryInfC{$Γ ⊢ φ₁ ⇒ φ₂$,}
\end{bprooftree}
}
\end{equation*}

\begin{equation*}
(\mathcal{D}_2)
\begin{bprooftree}
\AxiomC{$Γ ⊢ unshunt(φ₁ ⇒ φ₂) ∧ unshunt(φ₁ ⇒ φ₃)$}
\RightLabel{∧-proj₂}
\UnaryInfC{$Γ ⊢ unshunt(φ₁ ⇒ φ₃, n)$}
\RightLabel{By def.}
\UnaryInfC{$Γ ⊢ unshunt(φ₁ ⇒ φ₃, n-1)$}
\RightLabel{lem-unshunt}
\UnaryInfC{$Γ ⊢ φ₁ ⇒ φ₃$,}
\end{bprooftree}
\end{equation*}

@@

\begin{equation}
\label{eq:split-definition}
\mysplit(φ) =
\begin{cases}
unshunt(\mysplit(φ₁)) &\text{if }φ ≡ φ₁ ∧ φ₂\\
∧\ unshunt(φ₁ ⇒ \mysplit(φ₂)), &\\[2.5mm]
unshunt(¬ φ₁ ⇒ \mysplit(φ₂)),
  & \text{if }φ ≡ φ₁ ∨ φ₂\\[2.5mm]
unshunt(φ₁ ⇒ \mysplit(φ₂)),
  & \text{if }φ ≡ φ₁ ⇒ φ₂\\[2.5mm]
unshunt(φ₁ ⇒ \mysplit(φ₂))
  & \text{if }φ ≡ φ₁ ⇔ φ₂\\
∧\ unshunt(φ₂ ⇒ \mysplit(φ₁)),
  &\\[2.5mm]
unshunt(φ₁ ⇒ \mysplit(¬ φ₂)),
  & \text{if }φ ≡ ¬ (φ₁ ∧ φ₂)\\[2.5mm]
unshunt(\mysplit(¬ φ₁))
  & \text{if }φ ≡ ¬ (φ₁ ∨ φ₂)\\
∧\ unshunt(¬ φ₁ ⇒ \mysplit(¬ φ₂)),
  &\\[2.5mm]
unshunt(\mysplit(φ₁))
  & \text{if }φ ≡ ¬ (φ₁ ⇒ φ₂)\\
∧\ unshunt(φ₁ ⇒ \mysplit(¬ φ₂)),
  &\\[2.5mm]
unshunt(φ₁ ⇒ \mysplit(¬ φ₂))
  & \text{if }φ ≡ ¬ (φ₁ ⇔ φ₂)\\
∧\ unshunt(¬ φ₂ ⇒ \mysplit(φ₁)),
  &\\[2.5mm]
unshunt(\mysplit(φ₁)),
  & \text{if }φ ≡ ¬ ¬ φ₁\\
⊤,
  & \text{if }φ ≡ ¬ ⊥\\
⊥,
  & \text{if }φ ≡ ¬ ⊤.
\end{cases}
\end{equation}

@@

\begin{equation*}
(\mathcal{D})\hspace{3mm}
  \begin{bprooftree}
  \AxiomC{$Γ ⊢ \mysplit(φ,n)$}
  \RightLabel{By def.}
  \UnaryInfC{$Γ ⊢ unshunt(φ₁) ∧ unshunt(φ₁ ⇒ \mysplit(φ₂,n-1))$}
  \RightLabel{∧-proj₁}
  \UnaryInfC{$Γ ⊢ unshunt(φ₁)$}
  \RightLabel{lem-unshunt}
  \UnaryInfC{$Γ ⊢ φ₁$.}
  \end{bprooftree}
\end{equation*}

\begin{equation*}
  \begin{bprooftree}
  \AxiomC{$\mathcal{D}$}
  \UnaryInfC{Γ ⊢ φ₁}
  \AxiomC{$Γ ⊢ \mysplit(φ,n)$}
  \RightLabel{By def.}
  \UnaryInfC{$Γ ⊢ unshunt(φ₁) ∧ unshunt(φ₁ ⇒ \mysplit(φ₂,n-1))$}
  \RightLabel{$∧$-proj$₂$}
  \UnaryInfC{$Γ ⊢ unshunt(φ₁ ⇒ \mysplit(φ₂,n-1))$}
  \RightLabel{lem-unshunt}
  \UnaryInfC{$Γ ⊢ φ₁ ⇒ \mysplit(φ₂,n-1)$}
  \RightLabel{$⇒$-elim}
  \BinaryInfC{$Γ ⊢ \mysplit(φ₂,n-1)$}
  \RightLabel{\thmsplit}
  \UnaryInfC{$Γ ⊢ φ₂.$}
  \end{bprooftree}
\end{equation*}

@@

($φ ≡ φ₁ ⇒ φ₂$).
\begin{equation*}
  \begin{bprooftree}
  \AxiomC{}
  \RightLabel{assume φ₁}
  \UnaryInfC{$Γ , φ₁ ⊢ φ₁$}
  \AxiomC{$Γ ⊢ \mysplit(φ,n-1)$}
  \RightLabel{By def.}
  \UnaryInfC{$Γ ⊢ unshunt(φ₁ ⇒ \mysplit(φ₂,n-1))$}
  \RightLabel{lem-unshunt}
  \UnaryInfC{$Γ ⊢ φ₁ ⇒ \mysplit(φ₂,n-1)$}
  \RightLabel{weaken φ₁}
  \UnaryInfC{$Γ , φ₁ ⊢ φ₁ ⇒ \mysplit(φ₂,n-1)$}
  \RightLabel{⇒-elim}
  \BinaryInfC{$Γ , φ₁ ⊢ \mysplit(φ₂,n-1)$}
  \RightLabel{\thmsplit}
  \UnaryInfC{$Γ , φ₁ ⊢ φ₂$}
  \RightLabel{⇒-intro}
  \UnaryInfC{$Γ ⊢ φ₁ ⇒ φ₂$.}
  \end{bprooftree}
\end{equation*}
\vskip 3mm

@@

\begin{equation*}
(\mathcal{D})\hspace{3mm}
\begin{bprooftree}
\AxiomC{$Γ ⊢ \mysplit(φ, n)$}
\RightLabel{By def.}
\UnaryInfC{$Γ ⊢ unshunt(¬ φ₁) ∧ unshunt(¬ φ₁ ⇒ \mysplit(φ₂, n))$}
\RightLabel{∧-proj₁}
\UnaryInfC{$Γ ⊢ unshunt(¬ φ₁)$}
\RightLabel{lem-unshunt}
\UnaryInfC{$Γ ⊢ ¬ φ₁$,}
\end{bprooftree}
\end{equation*}

@@

  \AxiomC{$Γ ⊢ \mysplit(φ,n-1)$}
  \RightLabel{By def.}
  \UnaryInfC{$Γ ⊢ unshunt(¬ φ₁) ∧ unshunt(¬ φ₁ ⇒ \mysplit(φ₂,n-1))$}
  \RightLabel{∧-proj₂}
  \UnaryInfC{$Γ ⊢ unshunt(¬ φ₁ ⇒ \mysplit(φ₂,n-1))$}
  \RightLabel{lem-unshunt}
  \UnaryInfC{$Γ ⊢ ¬ φ₁ ⇒ \mysplit(φ₂,n-1)$}
  \RightLabel{⇒-elim}
  \BinaryInfC{$Γ ⊢ \mysplit(φ₂,n-1)$}
  \RightLabel{\thmsplit}
  \UnaryInfC{$Γ ⊢ ¬ φ₂.$}
  \end{bprooftree}
\end{equation*}
\end{proof}

@@

\begin{equation*}
  \begin{bprooftree}
  \AxiomC{$Γ ⊢ q$}
  \RightLabel{∨-intro₂}
  \UnaryInfC{$Γ ⊢ p ∨ q$}
  \RightLabel{∨-intro₁}
  \UnaryInfC{$Γ ⊢ (p ∨ q) ∨ r$}

  \AxiomC{$Γ ⊢ p$}
  \RightLabel{∨-intro₁}
  \UnaryInfC{$Γ ⊢ p ∨ q$}
  \RightLabel{∨-intro₁}
  \UnaryInfC{$Γ ⊢ (p ∨ q) ∨ r$}

  \RightLabel{∨-elim}
  \BinaryInfC{$Γ, q ∨ p ⊢ (p ∨ q) ∨ r$}
  \end{bprooftree}
\end{equation*}

\begin{equation*}
  \begin{bprooftree}
  \AxiomC{$Γ ⊢ r$}
  \RightLabel{∨-intro₂}
  \UnaryInfC{$Γ ⊢ (p ∨ q) ∨ r$}

  \AxiomC{$Γ, q ∨ p ⊢ (p ∨ q) ∨ r$}

  \RightLabel{∨-elim}
  \BinaryInfC{$Γ, r ∨ (q ∨ p) ⊢ (p ∨ q) ∨ r$}

  \RightLabel{⇒-intro}
  \UnaryInfC{$Γ ⊢ r ∨ (q ∨ p) ⇒ (p ∨ q) ∨ r$}

@@

In a disjunction, $φ = φ₁ ∨ φ₂ ∨ \cdots ∨ φₙ$, we say $ψ ∈_{∨} φ$,
if there is some $i = 1, \cdots, n$ such that $ψ = φᵢ$.
Then, we define the $rm_{∨}$ function to remove in a right-associative disjunction φ, those propositions ψ repeated in the formula.

@@

Now, we have removed redundancies in the disjunctions, we move forward and we remove redundancies at the level of the conjunction of the CNFs by applying the $rm_{∧}$ function. We define $ψ ∈_{∧} φ$ similar as we described above for disjunction. In a conjunction, $φ = φ₁ ∧ φ₂ ∧ \cdots ∧ φₙ$, we say
$ψ ∈_{∧} φ$, if there is some $i = 1, \cdots, n$ such that $ψ = φᵢ$.

@@

% subsection normalization (end)
% --------------------------------------------------------------------
\begin{lemma}[lem-rmPEM$_{∨}$]
  \label{lem:lem-rmPEM-or}
  \begin{align*}
  \label{eq:rmPEM-or}
    \begin{split}
    rmPEM_{∨}(φ) =
      \begin{cases}
        ⊤, &\text{if }φ ≡ ¬ φ₁ ∨ φ₂,\ φ₁ ∈_{∨} φ₂\\
        ⊤, &\text{if}φ ≡ ¬ φ₁ ∨ φ₂,\ rmPEM_{∨}(φ₂) ≡ ⊤\\
        ¬ φ₁ ∨ rmPEM_{∨}(φ₂), &\text{if }φ ≡ ¬ φ₁ ∨ φ₂\\
        ⊤, &\text{if }φ ≡ φ₁ ∨ φ₂,\ ¬φ₁ ∈_{∨} φ₂ \\
        ⊤, &\text{if }φ ≡ φ₁ ∨ φ₂,\ rmPEM_{∨}(φ₂) ≡ ⊤\\
        φ₁ ∨ rmPEM_{∨}(φ₂), &\text{if }φ ≡ φ₁ ∨ φ₂\\
        φ, &\text{otherwise.}
      \end{cases}
    \end{split}
  \end{align*}
\end{lemma}

\begin{lemma}[lem-rmPEM$_{∧∨}$]
  \label{lem:lem-rmPEM-and-or}
  \begin{equation*}
    \label{eq:rmPEM-and-or}
    rmPEM_{∧∨}(φ) =
    \begin{cases}
      rmPEM_{∨}(φ₁) ∧ rmPEM_{∨∧}(φ₂), &\text{ if }φ ≡ φ₁ ∧ φ₂\\
      rmPEM_{∨}(φ), &\text{otherwise.}
    \end{cases}
  \end{equation*}
\end{lemma}

\begin{lemma}[lem-rmBot$_{∧}$]
  \label{lem:lem-rmbot-and}
   $Γ ⊢ φ$ then $Γ ⊢ rmBot_{∧}(φ)$, where,
  \begin{equation*}
    \label{eq:rmBot-and}
    rmBot_{∧}(φ) =
    \begin{cases}
      ⊥, &\text{ if }φ ≡ ¬ φ₁ ∧ φ₂,\ φ₁ ∈_{∧} φ₂\\
      ⊥, &\text{ if }φ ≡ ¬ φ₁ ∧ φ₂,\ rmBot_{∧}(φ₂) ≡ ⊥\\
      ¬ φ₁ ∧ rmBot_{∧}(φ₂), &\text{if }φ ≡ ¬ φ₁ ∧ φ₂\\
      ⊥, &\text{ if }φ ≡ φ₁ ∧ φ₂,\ φ₁ ∈_{∧} φ₂\\
      ⊥, &\text{ if }φ ≡ φ₁ ∧ φ₂,\ rmBot_{∧}(φ₂) ≡ ⊥\\
      φ₁ ∧ rmBot_{∧}(φ₂), &\text{if }φ ≡ φ₁ ∧ φ₂\\
      φ, &\text{otherwise.}
    \end{cases}
  \end{equation*}
\end{lemma}

\begin{lemma}[lem-canon]
  \label{lem:lem-canon}
  \begin{align*}
  \begin{split}
  &canon(φ) =\\
  &\hspace{3mm}
  \begin{cases}
      canon(φⱼ), &\text{if } φ ≡ φ₁ ∧ φ₂,\ canon(φᵢ) ≡ ⊤, i,j = 1,2, i≠j\\
     ⊥, &\text{if } φ ≡ φ₁ ∧ φ₂,\ canon(φᵢ) ≡ ⊥, i = 1,2\\
     canon(φ₁) ∧ canon(φ₂), &\text{if } φ ≡ φ₁ ∧ φ₂\\
     ⊤, &\text{if } φ ≡ φ₁ ∨ φ₂,\ canon(φᵢ) ≡ ⊤, i = 1,2\\
     canon(φⱼ), &\text{if } φ ≡ φ₁ ∨ φ₂,\ canon(φᵢ) ≡ ⊥,  i,j = 1,2, i≠j\\
     canon(φ₁) ∨ canon(φ₂), &\text{if } φ ≡ φ₁ ∨ φ₂\\
     φ &\text{otherwise.}
    \end{cases}
  \end{split}
  \end{align*}
\end{lemma}

\begin{theorem}[thm-canonicalize] % (fold)
  \label{thm:thm-canonicalize}
  $Γ ⊢ φ$ then $Γ ⊢ canonicalize(φ)$, where,
  \begin{equation*}
    canonicalize =
     canon ∘ rmBot_{∧} ∘ rmPEM_{∧∨} ∘ redun ∘ rconj ∘ cnf.
  \end{equation*}
\end{theorem}

The \canonicalize rule acts in \Metis' \TSTP derivations in three main cases. First, introducing the
negation of a subgoal. Second, to simplify an intermediate derivation.
Third, introducing an axiom or a definition of the.
To justify the two first cases, we use the Theorem~\ref{thm:thm-canonicalize}. For the last case, we use the Lemma~\ref{lem:lem-nnf}.

% -------------------------------------------------------------------
\subsubsection{Simplify}
\label{sssec:simplify}

% subsubsection simplify (end)
% -------------------------------------------------------------------
